Ensemble Stability of Biological Networks

### 2014-09-16 paper.tex
move tables/figures to end and refactor into separate files

%------------------------------------------------------------------------------
% Beginning of paper.tex
%------------------------------------------------------------------------------
%
% AMS-LaTeX version 2 sample file for journals, based on amsart.cls.
%
%        ***     DO NOT USE THIS FILE AS A STARTER.      ***
%        ***  USE THE JOURNAL-SPECIFIC *.TEMPLATE FILE.  ***
%
% Replace amsart by the documentclass for the target journal, e.g., tran-l.
%
\documentclass{amsart}

%     If your article includes graphics, uncomment this command.
\usepackage{graphicx}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{xca}[theorem]{Exercise}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\numberwithin{equation}{section}

%    Absolute value notation
\newcommand{\abs}[1]{\lvert#1\rvert}

%    Blank box placeholder for figures (to avoid requiring any
%    particular graphics capabilities for printing this document).
\newcommand{\blankbox}[2]{%
  \parbox{\columnwidth}{\centering
%    Set fboxsep to 0 so that the actual size of the box will match the
%    given measurements more closely.
    \setlength{\fboxsep}{0pt}%
    \fbox{\raisebox{0pt}[#2]{\hspace{#1}}}%
  }%
}

% user added
\usepackage{longtable}
\usepackage{dot2texi}
\usepackage{tikz}
\usetikzlibrary{automata,shapes,arrows}
% todo notes see http://www.texample.net/tikz/examples/todo-notes/
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{hyperref}
\hypersetup{colorlinks=true,
linkcolor=[rgb]{.67 .27 .27},
citecolor=[rgb]{.09 .29 .54},
urlcolor=[rgb]{.09 .29 .54}}

% Set autoref text
% http://tex.stackexchange.com/a/36576/6784
\renewcommand*{\figureautorefname}{Fig.}
\renewcommand*{\equationautorefname}{Eq.}
\renewcommand*{\tableautorefname}{Table}
\renewcommand*{\sectionautorefname}{Section}

\def\tr{\mathrm{tr}}
\def\Path{\mathrm{Path}}
\def\hier{\mathrm{Hier}}
\def\Vertex{\mathrm{Vertex}}
\def\adj{\mathrm{adj}}

@@

For the case of continuous deterministic observables, the above process can be made more precise by associating a manifold to each observable, a directed graph to the interdependencies, and a vector field satisfying these interdependencies for each observable over the space determined by taking the product of the manifolds associated to the collection of observables \cite{Deville}. For example, if we have two observables $\{x_1,x_2\}$ where the domains in which they are quantified are given by manifolds $\{M_1,M_2\}$ such that $x_1 \in M_1 \equiv \mathbb{R}^1$ and $x_2 \in M_2 \equiv \mathbb{R}^1$, a directed graph $G_X$ describing the dependencies between these observables and vector field $V$ with components $\{F_1,F_2\}$ defined on $M_1 \otimes M_2 \equiv \mathbb{R}^2$ satisfying the dependencies determined by $G_X$. If the system under consideration has the graph
\begin{center}
\input{tex/examplehamdyngraph}
\end{center}
with adjacency matrix
$$
\adj(G_X) = \begin{pmatrix}
0 & 1 \\
1 & 1
\end{pmatrix}
$$
then for the system $X = \{G_X, M_X, F_X\}$, where $M_X \equiv \{M_1,M_2\}$ and $F_X \equiv \{F_1,F_2\}$ such that $F_1$ is constrained to be a function of $x_2$ and $F_2$ is a function of both $x_1$ and $x_2$ yielding the flow equations
\begin{align*}
\frac{dx_1}{dt} & = F_1(x_2)\\
\frac{dx_2}{dt} & = F_2(x_1,x_2)
\end{align*}

@@

In order to accommodate the possibility of uncertainty in our modelling,
we will generalize our notion of a system on a network to that of a
system with random parameters.  Again, this will involve three
steps. First, we provide a measure space $S$ which represents the
values over which our parameters can vary. Next, instead of a single
vector field, we consider a family of vector fields parameterized by
this space. That is to say, we introduce a mapping $V$ from $S$ to
sections of the tangent bundle of the product of manifolds. As before,
for each point $p \in S$, the vector field $V(p)$ must be consistent with
the system graph. Finally, we select a probability measure $\mu$ on
$S$ which represents our understanding of which values of the parameters
are likely. In accord with Bayesian statistics, we might revise this
distribution as data comes in or use it to estimate parameters and error
bars from experimental results.

Having done this, we are now in a position to do a probabilistic analysis.
Given some quantity $q$ characterizing our system (such as, for instance,
the time it would take a particle to travel between two points), this
quantity now becomes a random variable.  We may then study the average
value, variance, and the like.  Another question one can ask is that of
robustness, which is the conditional probability that, if some property
holds for a set of parameters, it will still hold if we make some random
perturbation about those parameter values.

This description can be formalized as a directed graph $G_X$ that describes the manner in which each of the variables depends upon one another given by an adjacency matrix $\adj(G_X)$ where
 \begin{displaymath}
   \adj(G_X)_{ij} = \left\{
     \begin{array}{ll}
       1, & F_i \hbox{ depends on } x_j\\
       0, & F_i \hbox{ does not depend on } x_j
     \end{array}
   \right.
\end{displaymath}
\todo{Define strongly connected components and connect to operads \cite{Vagner2014}}

@@

\begin{lemma}
Given a polynomial $p(z) = z^n + \sum_{k=1}^{n-1} s_k z^{n-k}$ and a real number
$\epsilon > 0$, there exists $\delta \in \mathbb{R}$ such that, if $p'(x) =
x^n + \sum_{k=1}^{n-1} {s'}_i z^i$ is a polynomial for which $|s_i - {s'}_i|
< \delta$ then to every root $r$ of $p$ there corresponds a root $r'$ of
$p'$ such that $|r-r'| < \epsilon$.
\end{lemma}

@@

We first consider the case where $p = x^n$.

@@

\begin{proof}
Since the mapping which sends the roots of a polynomial to its
coefficients has non-singular Jacobian when the discriminant is not
zero, if a stable polynomial has distinct roots, it does not lie on
the boundary.

@@

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.5\columnwidth]{fig/region2x2.pdf}
\caption{{\bf Stability conditions on coefficients of the characteristic polynomial for two component systems.} The regions correspond to all possible relationships between the invariants determined by the characteristic polynomial. Colors correspond to the regions given in the main text Green: $R_{2000}$, Red: $R_{1100}$, Yellow: $R_{0200}$, Magenta: $R_{0002}$, Blue: $R_{0020}$.}
\label{fig:region2x2}
\end{figure}

@@

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.5\columnwidth]{fig/region3x3.pdf}
\caption{{\bf Stability conditions on coefficients of the characteristic polynomial for three component systems.} Colors correspond to the following regions Green: $R_{3000}$, Red: $R_{1200}$, Yellow: $R_{1002}$, Blue: $R_{1020}$. This is the plane $s_3=1$.}
\label{fig:region3x3}
\end{figure}

@@

The characteristic polynomial $\chi(A)$ associated to $A$ can be factored according to the strongly connected components of the graph $G$. $$\frac{\chi(M)'(z)}{\chi(M)(z)} = \tr (zI - M)^{-1}$$ where, $\chi(M)(z) = \det (zI - M)$ is the chararacteristic polynomial of the matrix $M$ with $z$ a complex variable.  This formula is readily verified when $M$ is diagonal and has the nice property of being additive over decompositions of the matrix. Making the change of variables $x=\frac{1}{z}$, we have $$\tr (I - xM)^n = \sum_{n=0}^\infty x^n \sum_{a \in \Vertex(G)} \sum_{p \in \Path_n (G)(a,a)} F(p)$$
where $G$ is the graph of the matrix and $\Path_n (G;a,b)$ is the set of
paths of length $n$ from $a$ to $b$.  ; $F$ is the functor from this category to the multiplicative semigroup of the real numbers which corresponds to the matrix $M$.

@@

\newcommand{\specialcell}[2][c]{%
  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

\begin{table}[h]
\begin{tabular}{ c || c | c | c }
\hline
matrix & connectivity & \specialcell{probability of stability\\to perturbation} & probability of stability\\
\hline
  $\begin{pmatrix}
a & b \\
d & c
\end{pmatrix}$ & 4 & 0.62 & 0.25 \\
  $\begin{pmatrix}
a & b \\
d & 0
\end{pmatrix}$, $\begin{pmatrix}
0 & b \\
d & c
\end{pmatrix}$ & 3 & 0.5 & 0.25 \\
  $\begin{pmatrix}
a & 0 \\
d & c
\end{pmatrix}$, $\begin{pmatrix}
a & b \\
0 & c
\end{pmatrix}$ & 3 & 0.67 & 0.25 \\
$\begin{pmatrix}
a & 0 \\
0 & c
\end{pmatrix}$ & 2 & 0.5 & 0.25 \\
\end{tabular}
\caption{{\bf Probability of stability under resampling and \emph{a priori} stability for two component systems derived analytically}. All matrices not listed have $0$ probability of stability.}\label{tab:structstabmat}
\end{table}

@@

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.8\columnwidth]{fig/connectcycle3D3x3.pdf}
\caption{{\bf Plot of stability to perturbations versus number of cycles and connectivity for three component systems.} }
\label{fig:connectcycle3D3x3}
\end{figure}

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.8\columnwidth]{fig/stab3x3.pdf}
\caption{{\bf Plot of stability to perturbations versus connectivity for three component systems.} }
\label{fig:stab3x3}
\end{figure}

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.8\columnwidth]{fig/cycle3x3.pdf}
\caption{{\bf Plot of stability to perturbations versus number of cycles for three component systems.} }
\label{fig:cycle3x3}
\end{figure}

@@

\input{tex/tablenographs}

% \input{tex/tablewithgraphs}

\section{Appendix A}\label{sec:appA}

>>>>

\input{ams_header.tex}

@@

For the case of continuous deterministic observables, the above process can be made more precise by associating a manifold $M_i$ to each observable, a directed graph $G$ to the interdependencies, and a vector field $V \in T(\prod_i M_i)$ over the space determined by taking the product of the manifolds associated to the collection of observables, satisfying these interdependencies for each observable  \cite{Deville}. For example, if we have two observables $\{x_1,x_2\}$ where the domains in which they are quantified are given by manifolds $\{M_1,M_2\}$ such that $x_1 \in M_1 \equiv \mathbb{R}^1$ and $x_2 \in M_2 \equiv \mathbb{R}^1$, a directed graph $G_X$ describing the dependencies between these observables and vector field $V$ with components $\{F_1,F_2\}$ defined on $M_1 \otimes M_2 \equiv \mathbb{R}^2$ satisfying the dependencies determined by $G_X$. If the system under consideration has the graph given in \ref{fig:examplesystemmodules}A
% \begin{center}
% \input{tex/examplehamdyngraph}
% \end{center}
with adjacency matrix
$$
\adj(G_X) = \begin{pmatrix}
0 & 1 \\
1 & 1
\end{pmatrix}
$$
then for the system $X = \{G_X, M_X, F_X\}$, where $M_X \equiv \{M_1,M_2\}$ and $F_X \equiv \{F_1,F_2\}$ such that $F_1$ is constrained to be a function of $x_2$ and $F_2$ is a function of both $x_1$ and $x_2$ yielding the flow equations
\begin{align*}
\frac{dx_1}{dt} & = F_1(x_2)\\
\frac{dx_2}{dt} & = F_2(x_1,x_2)
\end{align*}

@@

In order to accommodate the possibility of uncertainty in our modelling,
we will generalize our notion of a system on a network to that of a
system with random parameters.  Again, this will involve three
steps. First, we provide a measure space $S, \mathcal{B}(S)$ which represents the
values over which our parameters can vary. Next, instead of a single
vector field, we consider a family of vector fields parameterized by
this space. That is to say, we introduce a mapping $V \colon S \to \Gamma T(\prod_i M_m)$.
(Here $\Gamma$ denoted the space of sections of the tangent bundle.)
As before, for each point $p \in S$, the vector field $V(p)$ must be consistent with
the system graph. Finally, we select a probability measure $\mu \colon
\mathcal{B}(S) \to [0,1]$ which represents our understanding of which values of the parameters
are likely. In accord with Bayesian statistics, we might revise this
distribution as data comes in or use it to estimate parameters and error
bars from experimental results.

Having done this, we are now in a position to do a probabilistic analysis.
Given some quantity $q$ characterizing our system, this quantity now
becomes a random variable.  This variable may be discrete of continuous
depending on the quantity under consideration.  For example, if the quantity
is the time it takes for a particle to travel between two points, we have
a real-valued random variable; if the quantity is the number of fixed
points, we have an integer-valued random variable; and if the quantity is
whether or not the system posesses a limit cycle, we have a binary random
variable.

Another question one can ask is that of robustness, which measures the
probability that, if some property holds for a set of parameters, it will
continue to hold if we make a random perturbation about those parameter values.
To define robustness, we will introduce, in addition to the probability
distribution $\mu$ described above, a family of probability distributions
$\mu' \colon S \times \mathcal{B}(S) \to [0,1]$.  Then, given a binary random
variable $q$, we we define its robustness as the following conditional proability:
\[
  \mathrm{Robust} (q) =
  \frac{\int_S d\mu(x) \int_S d\mu'(y) \mathbf{1}_q(x) \mathbf{1}_q(y)}
  {\int_S d\mu(x) \mathbf{1}_q(x)}
\]

This description can be formalized as a directed graph $G_X$ that describes the manner in which each of the variables depends upon one another given by an adjacency matrix $\adj(G_X)$ where
 \begin{displaymath}
   \adj(G_X)_{ij} = \left\{
     \begin{array}{ll}
       1, & F_i \hbox{ depends on } x_j\\
       0, & F_i \hbox{ does not depend on } x_j
     \end{array}
   \right.
\end{displaymath}
A graph can be decomposed into a natural collection of \emph{modules} called strongly connected components. A strongly connected component of a graph is a maximal subset of vertices where each vertex within the subset can be reached from any other \cite{Cormen2009}. The strongly connected components of some examples of three component systems are outlined in \ref{fig:scc}.

@@

\begin{lemma}
Given a polynomial $p(x) = z^n + \sum_{k=1}^{n-1} s_k x^{n-k}$ and a real number
$\epsilon > 0$, there exists $\delta > 0 \in \mathbb{R}$ such that, if $q(x) =
x^n + \sum_{k=1}^{n-1} \sigma_i x^i$ is a polynomial for which $|s_i - \sigma_i|
< \delta$ then one can order the roots of $p$ and $q$ such that $p(z) = \prod_{k=1}^n
(z - r_i)$ and $q(z) = \prod_{k=1}^n (z - \rho_i)$ and $| r_k - \rho_k | < \epsilon$.
\end{lemma}

@@

We first consider the case where $p(x) = x^n$.

@@

\begin{proof}
Suppose that none of the real parts of the roots of a polynomial $p$ equal
zero.  Let $\epsilon$ be the minimum of the absolute values of the real
parts of the roots of $p$.  Then, by continuity of the map which sends the
roots of a polynomial to its coefficients, there exists a $\delta > 0$
such that, if $q$ is any polynomial whose coefficients differ by less than
$\delta$ from the

@@

@@

@@

The characteristic polynomial $\chi(A)$ associated to $A$ can be factored
according to the strongly connected components of the graph $G$.
\[
\frac{\chi(M)'(z)}{\chi(M)(z)} = \tr (zI - M)^{-1}
\]
where, $\chi(M)(z) = \det (zI - M)$ is the chararacteristic polynomial of the
matrix $M$ with $z$ a complex variable.  This formula is readily verified when
$M$ is diagonal and has the nice property of being additive over decompositions
of the matrix. Making the change of variables $x=\frac{1}{z}$, we have
\[
\tr (I - xM)^n = \sum_{n=0}^\infty x^n \sum_{a \in \Vertex(G)} \sum_{p \in \Path_n (G)(a,a)} F(p)
\]
where $G$ is the graph of the matrix and $\Path_n (G;a,b)$ is the set of
paths of length $n$ from $a$ to $b$.  ; $F$ is the functor from this category
to the multiplicative semigroup of the real numbers which corresponds to the matrix $M$.

As an illustration of this formula, we will use it to compute the characteristic
polynomial of the matrix associated to the labelled graph in figure ??.  We begin
by noting that the closed paths of this graph can be described by the following
regular expressions:
\begin{align*}
\Path (1,1) &: (e + bd + bca)^* \\
\Path (2,2) &: (ae^*(be^*d)^*bc)^* \\
\Path (3,3) &: ((ca + d)e^*b)^* \\
\end{align*}
Since these regular expressions are non-redundant, we can obtain generating
functions for the paths by considering $a,b,c,d,e$ as non-commutative
variables and replacing the Kleene star $x^*$ by $(1 - x)^{-1}$:
\begin{align*}
\Path (1,1) &: (1 - e - bd - bca)^{-1} \\
\Path (2,2) &: (1 - a(1-e)^{-1}(1 - b(1-e)^{-1}d)^{-1}bc)^{-1} \\
\Path (3,3) &: (1 - (ca + d)(1 - e)^{-1}b)^{-1} \\
\end{align*}
To obtain the characteristic polynomial, we make the substitutions
$a \mapsto m_{12}/z, b \mapsto m_{31}/z, c \mapsto m_{23}/z, d \mapsto m_{13}/z,
e \mapsto m_{11}/z$ and simplify, using the fact that the matrix entries
commute:
\begin{align*}
\Path (1,1) &: \frac{z^3}
{z^3 - m_{11} z^2 - m_{13} m_{31} z - m_{12} m_{23} m_{31} } \\
\Path (2,2) &: \frac{z(z^2 - m_{11} z - m_{13} m_{31})}
{z^3 - m_{11} z^2 - m_{13} m_{31} z - m_{12} m_{23} m_{31} }  \\
\Path (3,3) &: \frac{z^2(z^2 - m_{11})}
{z^3 - m_{11} z^2 - m_{13} m_{31} z - m_{12} m_{23} m_{31} } \\
\end{align*}
Adding these three quantitites together, we obtain
\[
\frac{z(3 z^2 - 2 m_{11} z - m_{13} m_{31})}
{z^3 - m_{11} z^2 - m_{13} m_{31} z - m_{12} m_{23} m_{31} },
\]
which we recognize as $\chi'(z)/\chi(z)$, where
$\chi(z) = z^3−m_{11} z^2 - m_{13} m_{31} z - m_{12} m_{23} m_{31}$.
finally, we note that this polynomial is indeed the characteristic
polynomial of the matrix
\[
 \begin{matrix}
   m_{11} & m_{12} & m_{13} \\
   0      & 0      & m_{23} \\
   m_{31} & 0      & 0 \\
 \end{matrix}
\]
corresponding to our digraph.

@@

@@

@@

\newpage
\FloatBarrier

\section{Figure Legends}
\input{tex/figures}

\newpage
\FloatBarrier

\section{Tables}
\input{tex/table2x2stab}
\input{tex/tablenographs}
% \input{tex/tablewithgraphs}
